Predicting the conformational preferences of proteins using a physics-based free energy method

### 2013-04-24 Paper/paper.tex
correction_24thApril

\author{Arijit~Roy}
\author{Alberto~Perez}
\author{Ken~A.~Dill}
\author{Justin~L.~MacCallum}
\affil{Laufer Center for Physical and Quantitative Biology\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

@@

@@

Protein conformation and associated conformational changes are important in biology. The mechanism
of enzymes, the function of structural proteins, and the binding of proteins to DNA or other
proteins all depend on protein conformations and the relative free energies of different
conformational states. Proteins can undergo conformational changes for a variety of reasons,
including ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007},
\cite{Jorgensen2004}. The process of protein folding also depends on the free energies of different
conformational states. The native structure is at a free energy minimum \cite{Anfinsen1973}, while
the kinetics are determined by the free energies of various on- and off-pathway intermediates.

However, the theoretical calculation of conformational free energy changes is difficult when the
states of interest are very different \cite{Meirovitch2007}. In these cases, the timescales involved
in such conformational transitions may be too long to simulate directly. For example, in protein
folding the free energy landscape is rough \cite{Dill1997}, \cite{Dill2008}, and the molecule can
become trapped in an energy well during conformational sampling. Even with advanced methods like
replica exchange molecular dynamics, the three dimensional structure may be trapped in a local
minima for a considerable time. Sometimes this problem can be circumvented by defining a pathway or
reaction coordinate connecting the states of interest. The free energy along this reaction
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977},
targeted MD \cite{}, or thermodynamic integration \cite{Tironi1994}. However, accurately sampling
this reaction coordinate may still be difficult and require a large computationl effort. In the case
of protein folding, it is often difficult to even devise a meaningful reaction coordinate that could
connect the unfolded and folded states efficiently \cite{}.

The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Meirovitch2007,Ytreberg2006,Zheng2008}. In recent years, methods such as the reference
system method \cite{Ytreberg2006}, decativated morphing \cite{Park2008}, orthogonal space random
walk \cite{Zheng2008}, and the confinement method \cite{Tyka2006,Cecchini2009} have been used to
calculate free energy differences.

In this work we explore confinement method\cite{Tyka2006,Cecchini2009}, which relies on the fact
that the free energy difference is a state function and thus is independent of the sampling path.
This approach uses a thermodynamic cycle \ref{fig:method} where first protein conformational entropy
of the two end-states is frozen out by applying a series of restraints to the system. Next, the free
energy difference between the two highly-restrained harmocic systems is obtained using a normal mode
calculation.

Previously, this method was applied to calculate the conformational free energy of a 16 amino acid
residue peptide, known as BHP.  We have first reproduced that known result. Next, we have applied
this method to larger proteins in two different case studies. In the first, we attempt to pick the
most native like structures from a set of models produced during the Critical Assessment of
Structure Prediction (CASP) \cite{} experiment. In the second case study, we examine the
conformational preferences of a series of chameleon proteins designed by Orban and co-workers
\cite{} that can switch between two completely different folds with only small changes in amino acid
sequence.

@@

\subsection{Our implementation of the confiment method reproduces results from the literature}

As a first step, we performed several control experiments to verify that our implementation of the
confinement method produces results compatible with previous calculations reported in the
literature.

The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from protein G,
known as BHP \cite{Cecchini2009}. We calculated the free energy difference between two different
conformations of the peptide: (1) the native conformation, called bhp1, with a two-stranded
$\beta$-sheet; and (2) a conformation, called bhp3, which has a three-stranded $\beta$-sheet.
Analysis of long (4 $\mu$s) equilibrium simulations \cite{Cecchini2009,Krivov2004}  shows that bhp1
is the more favorable configuration by 1.8 kcal/mol. Using the confinement method, we obtain a value
1.7 kcal/mol, which is in good agreement with the equilibrium simulations and with previous
calculations using the confinement method \cite{Cecchini2009}.

\subsection{The confiment method predicts that experimental structures have lower free energies than
computer-generated models}

Previous applications of the confiment method have focused on relatively short peptides, up to 17
residues in length. In this work we apply the confinement method to larger proteins. As a first
test, we calculated the free energies of models submitted by predictors during the CASP9 experiment
(often referred to as decoys) and compared them with the calculated free energy of the corresponding
experimentally determined structure from the Protein Data Bank (PDB). As
Table~\ref{table:casp_control} shows, in 5 out of 6 cases, the confinement method assigns a lower
free energy to the experimentally determined structure than to any of the decoys---as would be
expected. Although differentiating between experimentally determined structures and computer
generated predictions is not a stringent test of a scoring method---for example, examining the
residue-residue packing can easily distinguish between the two [ref Rosetta Holes]---the results
none the less serve as a useful "sanity check".

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
    determined structure than to computer-generated predictions. For each target, we examined as
    many as five predictions submitted by CASP participants. We report the free energy difference
    between the most favorable decoy and the experimentally determined structure. Positive
    $\Delta\Delta G$ values indicate that the experimental structure is predicted to be more
favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}
\end{table}

\subsection{The confinement method correctly predicts the structural preferences of chameleon
sequences}

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). Sequences that adopt a
mixed alpha/beta structure similar to Protein G are denoted as ``GB'', while sequences that form a
three-helix bundle are denoted as ``GA''. We label the two folds as $3 \alpha$ and $4 \beta +
\alpha$. The GA sequences are observed to fold into the $3 \alpha$ fold and the GB sequences fold
into the $4 \beta + \alpha$ fold. One pair of sequences (GA88/GB88) are 88 percent identical in
sequence and differ in seven positions. Another pair (GA95/GB95) are 95 percent identical and differ
in three positions. The last pair (GA98/GB98) differ only in a single tyrosine to alanine mutation.
The fact such small changes in sequence can lead to such dramatic changes in structure is rather
remarkable. Accurately predicting the structural preferences of these structures presents a serious
challenge for computational methods \cite{Alexander2007}$-$\cite{Shortle20009}.

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all six sequences (data not shown). This is, however, not
surprising. It is well known [CITE] that it is easy to distinguish computational models from native
structures. To avoid this potential problem, we instead computed relative free energy of two
different computer generated models for each sequence. One model is based on the GA structure and
the other on the GB structure (see Supporting Information for details on the modeling procedure).
This is a much more realistic test of the confinement method's ability to accurately calculate
relative free energies.

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all five sequences. One hypothesis for fold switching of
chameleon sequences is that the structural transitions require states with diminished stability
[ref]. It is believed that if the free energy of the native state and the alternative state are
within a range of 5 Kcal/mol, then it is possible for the native state to be destabilized relative
to the alternative state with only small changes in sequence. The stability of the native state can
decrease for a number reason ranging from chemical modification, breaking of disulphide bonds or
mutations---as is the case here. The calculated free energy differences range from around 3.5 to 5.0
kcal/mol, which is consistent with the above hypothesis \cite{He2008}, \cite{{Alexander2009}}.

\subsection{Per-residue free energy calculation can identify the mechanistic detail behind
conformational preferences}

To understand the mechanism behind the conformational preferences of the different protein
sequences, we decomposed the calculated free energy into per-residue contributions as described in
the method section. This helps us to identify important residues that stabilize a particular
conformation. The per residue free energy, $\Delta \Delta G ((4 \beta + \alpha$) - ($3 \alpha))$ is
shown in Figure~\ref{fig:orban_full}(F). In this plot, a negative peak indicates that a residue
favors the $4 \beta + \alpha$ conformation and a positive peak indicates that the residue favors the
$3 \alpha$ structure.

The GA and GB sequences are nearly identical, so there are some common features observed for all
sequences. Roles of some important residues in stabilizing either $4 \beta + \alpha$ or $3 \alpha$
conformation and possible reason for such conformational preferences are discussed in details in
Figure~\ref{fig:orban_full} and Table~\ref{table:orban_perresidue}. The experimental observations
classified the protein into two parts: Amino acids 9-51 are fully structured in both the folds,
where as residues 1-8 and 52-56 are unstructured in $3 \alpha$, but form $\beta$ strand in $4 \beta
+ \alpha$ structure. Most of the amino acid residues in the region 1-9 favor $4 \beta + \alpha$
structure as can be seen from the plot Figure~\ref{fig:orban_full}(F). We can explain most of the
major peaks in this plot. For example,  first major peak appears at residue 7. Here, the hydrophobic
Leu-7 stabilizes the $4 \beta + \alpha$ conformation as its sidechain is oriented towards the
protein hydrophobic core, whereas in $3 \alpha$ structure it exposed to the solvent
(Figure~\ref{fig:orban_full}(A)). Thus, we hypothesise that a mutation of Leu-7 to a hydrophilic
residue would stabilize the $3 \alpha$ fold. The conformational preferences of most of the other
large peaks can be similarly explained by hydrogen bonding, salt bridge formation, and solvent
exposure of hydrophobic residues. They are listed in Figure~\ref{fig:orban_full} and
Table~\ref{table:orban_perresidue}.

We will now discuss the differences between the $3 \alpha$ and $4 \beta + \alpha$ folds due to the
specific differences between the GA and GB sequences. The differences are in positions 20, 30, and
45. GA95 has Leu-20, Ile-30, and Leu-45; GB95 has Ala-20, Phe-30, and Tyr-45. In the $3 \alpha$
fold, residues 20 and 30 are involved in interactions with residue 49. The peak at position 49
stabilizes the GA sequence more than that of GB sequence. In the $3 \alpha$ fold, Ile-49 has
hydrophobic interactions with Leu-20 and Ile-30 with GA sequence, whereas with GB sequence it has
such interactions with smaller Ala-20 (Figure~\ref{fig:orban_full}(I). Also, in the $3 \alpha$ fold
with the GB sequence, Phe-30 being a larger group can not be completely accommodated in the
hydrophobic core (Figure~\ref{fig:orban_full}(I)) and therefore, partially exposed to the protein
surface. There is an interesting difference for residue at position 45 as well. In GB95 sequence and
$4 \beta + \alpha$ fold there is H-bonding between Tyr-45 and Asp-47
(Figure~\ref{fig:orban_full}(K)), which is absent in the $3 \alpha$ conformation. On the other hand
with GA sequence, this H-bond is absent for both the conformation as there is a Leu at position 45
(Figure~\ref{fig:orban_full}(J)).

\begin{table}
\begin{center}
\caption{Analysis of the per-residue free energy decomposition of GA95 and GB95 sequence reveal that certain residues prefer
either $3 \alpha$  or $4 \beta + \alpha$ structure. These observations and the reason behind such conformational preferences are listed in this
table.}
\label{table:orban_perresidue}
\begin{tabular}{| l | l | l |}
\hline
    Information from per                                 & What is happening in           & What is happening in                      \\
    residue plot                                         & $3 \alpha$ conformation        & $4 \beta + \alpha$  conformation          \\ \hline
    Leu-7 favors                                         & hydrophobic residue is         & oriented towards the                      \\
    $4 \beta + \alpha$ (Figure~\ref{fig:orban_full}(A))  & exposed to the solvent         & hydrophobic core                          \\ \hline
    Gln-11 stabilizes $3 \alpha$                   & Gln-11 has H-bond                         & no such H-bond                           \\
    (Figure~\ref{fig:orban_full}(B))           &  with Glu-15                              &                                           \\ \hline
    Glu-14 favors $4 \beta + \alpha$           & hydrophilic residue partially  & completely exposed                                    \\
    (Figure~\ref{fig:orban_full}(E))           & exposed to the solvent         & to the solvent                                        \\ \hline
    Ala-26 favors  $4 \beta + \alpha$          & hydrophobic residue            & oriented towards the                                   \\
     (Figure~\ref{fig:orban_full}(C))          & oriented towards surface       & hydrophobic core                                       \\ \hline
    Ile-49 favors $3 \alpha$                   & hydrophobic interaction        & sidechain exposed                                       \\
    (Figure~\ref{fig:orban_full}(H) and (I))   & inside the protein             & to the solvent                                          \\ \hline
    Ile-49 favors GA more                      & more hydrophobic               &  unchanged                                              \\
    than GB                                    & interactions with residues     &                                                         \\
    (Figure~\ref{fig:orban_full}(H) and (I))  &  at 20, 30 in GA                &                                                         \\ \hline
    Val-54 favors $4 \beta + \alpha$           & hydrophobic residue            & hydrophobic interaction                                \\
    Figure~\ref{fig:orban_full}(D))             & exposed to the solvent        & inside the protein                                      \\ \hline
    Glu-56 favors $3 \alpha$                   & hydrophilic residue partially  & completely exposed                                    \\
    (Figure~\ref{fig:orban_full}(G))            & exposed to the solvent         & to the solvent                                        \\ \hline
    Phe-30 in GB sequence                      & exposed to the solvent         & oriented towards the                                  \\
    stabilizes $4 \beta + \alpha$                  & in protein surface             & hydrophobic core                                      \\ \hline
    Tyr-45 in GB favors                      & No H-bonding                   & H-bond with Asp-47                                    \\
    $4 \beta + \alpha$ (Figure~\ref{fig:orban_full}(K))&                 &                                                       \\ \hline
    Asp-47 in GA sequence                      & H-bond with Lys-50             & No H-bond, No Tyr-45                                  \\
    favors $3 \alpha$                          &                         & in GA sequence                                         \\ \hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}
\includegraphics[width=3.5 in,height=4 in]{orban.pdf}
\label{fig:orban}
\caption{Confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

\begin{figure}
\includegraphics[width=6.4 in,height=5.3 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Sidechain orientation of some important residues which stabilize/destabilize either of the $4 \beta + \alpha$ or $3 \alpha$ conformation.
The details are in Table~\ref{table:orban_perresidue} and the main text.
A. Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Ala-26
D. Val-54.
E. Glu-14.
F. plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak
stabilize $4 \beta + \alpha$.
G. Glu-56 .
H. sidechain orientation of Hydrophic Ile-49, along with Leu-20 and Ile-30 with GA95 sequence
I. sidechain orientation of Hydrophic Ile-49, along with Ala-20 and Phe-30 with GB95 sequence
J. Tyr-45 forms H-bond with Asp-47 with GB95 sequence and $4 \beta + \alpha$ conformation
K. No H-bond possible between Leu-45 and Asp-47 with GB95 sequence and $4 \beta + \alpha$ conformation .}
\end{figure}

\subsection{We have applied the confinement method to structure prediction}

We have applied the confinement method as a ``meta-predictor'' for structure prediction. Here, the
task is to correctly identify the most correct models out of a set of ``decoys'' generated by
different methods during the Critical Assessment of Structure Prediction (CASP) experiment. CASP is
a blind test in which different groups apply methods to predict the 3-dimensional structure of
proteins from their sequences. This is done with a either a three week (for humans) or three day
(for fully automated servers) time limit on each target. Each group is allowed to submit five
possible structures, which they are supposed to rank from best to worst. We have performed two
experiments centered on CASP.

In the first experiment, we examined several targets and tried to rank-order the predictions
produced by other groups. In some cases the structures all came from the same group, in others the
structures were drawn from several different groups. The goal of this experiment is to determine if
the physics-based confinement method can correctly identify more native-like structures as having
low free energies.

@@

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total
Score (GDT-TS) \cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can
be understood roughly as the percentage of residues that are correctly positioned in the model
(higher is better, range 0 to 100).

We have chosen the initial models on which to test the methods based on server groups that have
traditionally done well in past CASP events.

@@

Our next test was between models that are produced with different methodologies. We do this by
looking at some of the best CASP submissions from different research groups. For this purpose we
have chosen the x-ray crystallographic structure of fas apoptosis inhibitory protein molecule whose
pdb id and CASP target codes are 3MX7 and T0540 respectively. This protein contain 90 amino acid
residues with 8 beta strands. We have chosen best models from groups "LTB" and "Mufold" for our
analysis, which are labelled as Model 1 and Model 2 in Figure~\ref{fig:T0540}. As presented in
Figure~\ref{fig:T0540} we found that this method is able to correctly order the models which matches
with the GDT\_TS score.

>>>>

\author{Arijit~Roy}
\author{Alberto~Perez}
\author{Ken~A.~Dill}
\author{Justin~L.~MacCallum}
\affil{Laufer Center for Physical and Quantitative Biology\\
    and Departments of Physics and Chemistry\\
    Stony Brook University\\
    Stony Brook, NY 11794-5252.}

@@

\Ken{[Guys - I recommend we get a better name than confinement.  It's totally uninformative.  In my view, better would be `confine-and-configure' or something.  
Please give it some thought.]}

@@

In some problems of protein science, you want to know the relative stability of a protein's conformation $A$ compared to it's 
conformation $B$.  We call this the \emph{difference free energy} (DFE). For example, in allosteric mechanisms, a protein often 
adopts one conformation when a ligand is bound and another conformation when no ligand is bound \cite{Elber2007}. To understand the mechanism requires knowledge 
of the relative free energies of the protein conformations.  Another example is in protein structure prediction, for example as practiced in 
the CASP competition \cite{Moult2011}.  If you are using some computational model and have predicted two putative native structures, $A$ 
and $B$, you want to compute which has the lower free energy, in order to know which is the more native-like.  In a third example, you may 
want to know whether mutating a few amino acids in a protein could cause the protein to switch from one stable conformation to another, because 
that could have important consequences for biological mechanisms and disease.  And fourth, sometimes binding a ligand induces a protein from 
one conformation to another.  A quantitative understanding of such \emph{induced-fit} situations requires knowledge of the difference free 
energy of the protein in the two states. In all these cases, it would be useful to have a computational method that can be used efficiently 
with atomically detailed physical forcefield models to compute the free energy differences between different given protein conformations.

Methods like thermodynamic integration \cite{Tironi1994},\cite{Meirovitch2007} which are successful in alchemical free energy calculation
may not be as useful for calculation of conformational free energy.
One widely explored strategy is to use molecular dynamics simulations along some putative reaction coordinate pathway from conformation $A$ 
to $B$ \cite{Elber2007,West2007,Haas2007,Jonsson1998,E2007,Dellago2002,Cheng2006}. \Ken{we need 
many refs here: cite Elber's milestoning, Jhih-Wei Chu, and the many other path-sampling approaches.]} The free energy along this reaction 
coordinate can then be determined using methods such as umbrella sampling \cite{Torrie1977}. Such approaches have several limitations.  
First, it is necessary to know a reaction pathway from $A$ to $B$. If conformations $A$ and $B$ are quite different, then it can be 
challenging to find such paths. Second, these methods 
are computationally slow. To get an accurate estimate of the total free energy difference $\Delta G = G_B - G_A$ requires accurate determinations 
of the many small free energy differences $A \rightarrow 1 \rightarrow 2 \ldots \rightarrow B$, so each microstep requires substantial amounts 
of sampling.  And third, these methods are prone to large errors, because the pathway error is a sum of many errors among the many microsteps.

\Arijit{Will it be useful to add methods like PMF or replica exchange based free energy methods ?}

\Ken{[Are all these methods below ones that don't involve paths?  Also, are there other key methods we should cite?  We need a good scholarly 
list of all the relevant works.]} 
The calculation of protein conformational free energy has been successfully attempted by a number of
groups \cite{Shell2010,Ytreberg2006,Zheng2008,Spichty2010,Strajbl2000,Park2008,Tyka2006,Cecchini2009,Ovchinnikov2013}. 
Some of these methods like reference system method \cite{Ytreberg2006}, deactivated morphing \cite{Park2008}, confinement 
method \cite{Tyka2006,Cecchini2009,Ovchinnikov2013} take an alternative strategy for computing difference free energies 
that does not require knowing a pathway from $A$ to $B$.  
\Ken{[please add the several new Karplus papers from 2012-2013.]}.

Here, we adopt the confinement method of Tyka et al \cite{Tyka2006} and Cecchini et al.\cite{Cecchini2009}. This method 
also resembles the `confine-and-release' method \cite{Mobley2007}, \cite{Mobley2012}, \cite{Mobley2006}.  
\Ken{[cite mobley's jctc 07 (ref 216 on our labsite), and mobley's recent j chem phys review, 
and mobley (ref 199) j chem phys, and others that are relevant.]}. The method is based on the thermodynamic cycle showns in 
Figure \ref{fig:method}.  To start, conformation $A$ is an ensemble.  We reduce ensemble $A$ to one particular microstate $A^\ast$ 
in that ensemble by applying restraints in an MD simulation.  In the same way, we reduce conformational ensemble $B$ to one of its 
microstates, $B^\ast$.  Then, we simply compute the free energy difference between the microstates 
as $\Delta G_{A^\ast, B^\ast} = \Delta H_{A^\ast, B^\ast}$ (which is just an enthalpy difference since there is no conformational 
entropy difference between the microstates).  $\Delta H_{A^\ast, B^\ast}$ is readily calculated using normal mode analysis. In 
this way, we can calculate the free energy difference between the two end states without needing to define a physical path or reaction 
coordinate connecting the two states.

\Ken{For this fig, let's make A and A* green and B and B* red or something, to distinguish them visually.  Let's also blur the ensemble more relative to the microstate, so it's clearer the distinction.]}

Previously, the confinement method has been validated on small model peptides \cite{Tyka2006,Cecchini2009}. Here, we show that the confinement 
method can also be applied successfully to larger systems, such as the series of chameleon proteins designed by Orban and co-workers
\cite{Alexander2007,He2008,Alexander2009,He2012}, which can switch between two completely different folds with only few-amino-acid changes 
in sequence.  We also show that our computed difference free energies are useful in evaluating the quality of CASP target protein predictions.  
This could be of significant value for ultimately improving the energetics in protein-structure models, not just the structures.  And, 
finally, we show that we can approximately decompose difference free energies into individual amino acid components.  This offers the 
opportunity for diagnosing the structural basis for difference free energies, which can be useful for interpreting biological mechanisms.

@@

\subsection{The confinement method succeeds at some basic consistency checks}

First, we validated that our implementation of the confinement method produces results compatible with previous calculations reported in the 
literature.  The method has previously been applied to a 16 amino acid residue $\beta$-hairpin from protein G, known as 
BHP \cite{Cecchini2009}. We calculated the free energy difference between the native conformation (called bhp1), which has 
a two-stranded $\beta$-sheet and a three-stranded $\beta$-sheet called bhp3.  Our confinement calculation shows that bhp1 
is more stable by 1.7 kcal/mol, consistent with 4 $\mu$s equilibrium simulations showing that bhp1 is more favorable configuration 
by 1.8 kcal/mol, and in agreement with previous calculations \cite{Cecchini2009}.

Second, we looked at 6 target proteins from the CASP9 experiment.  For each target, we examined up to 5 submitted models.  We 
computed the difference free energy between the true native and the best model.  Table~\ref{table:casp_control} shows that, in 5 
out of 6 cases, the confinement method assigns a lower free energy to the experimentally determined structure than to any of the 
decoys.  Other well-known discriminators  can also successfully make this recognition \cite{Sheffler2009}; 
 it is just an independent useful validation that the confinement calculations make sense.

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
    determined structure than to computer-generated predictions. For each target, we examined as
    many as five predictions submitted by CASP participants. We report the free energy difference
    between the most favorable decoy and the experimentally determined structure. Positive
    $\Delta\Delta G$ values indicate that the experimental structure is predicted to be more
favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta \Delta G = G_{best~decoy} - G_{native}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}
\end{table}

\Ken{[ I think a better notation is $\Delta \Delta G = G_{best~decoy} - G_{native}$, rather than having the arrow.]}

\subsection{The confinement method correctly predicts the structures of chameleon sequences}

We also tested the confinement method predictions for difference free energies on the chameleon sequences of Orban 
et al. \cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}; these 
are instances in which two highly similar sequences fold into remarkably different structures. Orban and co-workers have designed a 
protein-G-like sequence of 56-residues that
is marginally stable in one of two possible folds. By mutating key residues in this sequence they
are able to stabilize one fold or the other (see Figure~\ref{fig:orban}). We call sequences ``$\beta$'' that fold into 
the protein-G-like that fold into a $4 \beta + \alpha$ fold, and we call sequences ``$\alpha$'' that fold into $3 \alpha$ helical folds.  One pair 
of sequences (GA88/GB88) is 88 percent identical in sequence, differing in seven positions. Another pair (GA95/GB95) is 95 percent identical, 
differing in three positions. The third pair (GA98/GB98) differs in only a single tyrosine-to-alanine mutation. Accurately predicting the 
structural preferences of these structures presents a serious challenge for computational methods 
\cite{Alexander2007,He2008,Alexander2009,Bryan2010,He2012,Shortle2009}.

\begin{figure}
\includegraphics[width=5.0in]{orban.pdf}
\label{fig:orban}
\caption{Confinement method correctly predicts the structural preferences of four chameleon
sequences. The top part of the figure represent four sequences used in this study along with the
protein data bank id. Each sequence adopts either a three-helix bundle fold (denoted $\alpha$) or
$4 \beta + \alpha$ fold (denoted $\beta$). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all four sequences (data not shown). This is, however, not
surprising. It is well known [CITE]\Arijit{still searching this citation} that it is easy to distinguish 
computational models from native structures. To avoid this potential problem, we instead computed relative free energy of two
different computer generated models for each sequence. One model is based on the $\alpha$ structure and
the other on the $\beta$ structure (see Supporting Information for details on the modeling procedure).
This is a much more realistic test of the confinement method's ability to accurately calculate
relative free energies.

\Ken{[ It's not clear what this test is and why it's better, and why there are only 5 sequences.  Please clarify.]}
\Arijit{ Now there are 4 sequences as they have the experimentally available structure. The fifth sequence is discussed 
at the end of the paragraph.}

Figure~\ref{fig:orban} shows that the confinement method identifies the correct structure for all four sequences. One hypothesis for fold switching of
chameleon sequences is that the structural transitions require states with diminished stability
\cite{Bryan2010}. It is believed that if the free energy of the native state and the alternative state are
within a range of 5 kcal/mol, then it is possible for the native state to be destabilized relative
to the alternative state with only small changes in sequence. The stability of the native state can
decrease for a number reason ranging from chemical modification, breaking of disulphide bonds or
mutations---as is the case here. The calculated free energy differences range from around 3.5 to 5.0
kcal/mol, which is consistent with the above hypothesis \cite{He2008,Alexander2009,Bryan2010}. 
In a more recent study\cite{He2012}, the amino acid residue at position 45 (Tyr for $\beta$ and Leu for $\alpha$) was
found to be important for switching between  $\alpha$ and $\beta$ conformation. This inspired us to 
introduce another mutation at 45 position. For our future reference this Tyr-45-Ala mutation of GB95 sequence will
be known as GB98. Interestingly, with this mutation, the equilibrium shifts to $\alpha$ conformation as it gains 3.84 Kcal/Mol 
stability compare to $\beta$ conformation and indicate the importance of Tyr-45 residue. It will be interesting to 
know the actutal experimental structure which can validate our result.

\subsection{Per-residue free energy calculation can identify the mechanistic detail behind
conformational preferences} \label{sec:orban_per_residue}

To understand the mechanism behind the conformational preferences of the different protein
sequences, we decomposed the calculated free energy into per-residue contributions in
an approximate way \cite{Tyka2006}. For this purpose, the confinement free energy, $\Delta G_{A,A*}$ and
$\Delta G_{B,B*}$ of each residue is calculated in the usual numerical way as described
the method section. We call this method approximate as we ignore the contribution from the normal 
mode or quasiharmonic analysis. Instead, decomp module of amber is used to calculate the internal energy 
of each residue from the final restrained trajectory.

\Ken{[Arijit -- I didn't find this methods calculation.  We should at least put in a sentence or two here summarizing it.]}  
This helps us to identify important residues that stabilize a particular
conformation. The per residue free energy, $\Delta \Delta G ((\beta) - (\alpha))$ is
shown in Figure~\ref{fig:perresidue_orban}. 
In this plot, residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure, and white residues
have no preference.

\begin{figure}
\begin{center}
\includegraphics[width=4.5in]{delta_g.pdf}
\end{center}
\caption{The $\alpha$ and $\beta$ conformation associated with GA95 and GB95 sequences are colored according to the relative per residue
free energy value. In this figure, residues colored in blue favor the $\alpha$ structure, residues in red favor the $\beta$ structure, and 
white residues have no preference. Some of the mutated residues are also labelled.}
\label{fig:perresidue_orban}
\end{figure}

Although the overall free energy difference between the two structures is small (within 3-5 Kcal/Mol), the individual residues 
can show marked preferences for being in either $\alpha$ or $\beta$
conformation. Such differences can be undestood by looking in detail at the local environemnt for
those resiudes. For example, the region around residue 7 forms a random coil in the $\alpha$
structure, whereas it is forming a beta sheet structure in the $\beta$ structure. These residues
strongly favour the locally well packed and hydrogen bonded environment found in the $\beta$ sheet.
Overall, favoring $\alpha$ or $\beta$ structure is a delicate balance, were the relative global free
energy difference is small and the contributions of different per residue tendencies balance out. It
is therefore very likely that by changing key resiude preferences the global fold preferences can be
changed.

Three residues are mutated between GA95 and GB95, at positions 20, 30, and 45. In GA95, these
residues (L20, I30 and L45) stabilize the $\alpha$ structure. In GB95, two of these residues (F30
and Y45) favor the $\beta$ structure, because they have large solvent exposed surface areas in the
$\alpha$ structure, but are more buried in the $\beta$ structure. Additionally Y45 forms a hydrogen
bond with D47 in the $\beta$ structure. On the other hand, residue A20 from GB95 still favors the
$\alpha$ structure, although not as strongly as L20 in GA95.

The $\alpha$ and $\beta$ sequences are nearly identical, so there are some common features observed for all
sequences. Roles of some important residues in stabilizing either the $\alpha$ or $\beta$
conformation and possible reasons for such conformational preferences are summerized in 
supporting information Figure S1. The experimental observations
classified the protein into two parts: Amino acids 9-51 are fully structured in both folds, whereas
residues 1-8 and 52-56 are unstructured in the $\alpha$ fold, but form a $\beta$-strand in the
$\beta$ fold. Most of the amino acid residues in the region 1-9 have negative per-residue free
energies, which means that these residues favor the $\beta$ structure.

In addition to the direct effects of the mutation, there are also indirect effects due to small
perturbations due to the mutations. For example, the L20A mutation causes a slight repacking around
residue 20. This causes large changes in the per-residue free energies of nearby residues T25 and
A26. However, these differences largely cancel and the overall effect of the L20A mutation is weaker
than either the I30F or L45Y mutations.

Such per-residue free energy decompositions allow us to understand the driving forces behind protein
folding and conformational change and may be useful for designing proteins with specific structures
and functions.

\subsection{Applcation of confinement method for structure prediction}

We applied the confinement method as a ``meta-predictor'' for structure prediction. Here, the
task is to correctly identify the most accurate models out of a set of ``decoys'' generated by
different methods during the Critical Assessment of Structure Prediction (CASP) experiment. CASP is
a blind test in which different groups apply methods to predict the 3-dimensional structure of
proteins from their sequences. Each group is allowed to submit five
possible structures, which they are supposed to rank from best to worst. We have performed two
experiments centered on CASP.

In the first experiment, we examined several targets and tried to rank-order the predictions
made by the same group (same methodology used to generate the models), in others the
structures were drawn from several different groups (different methodology used). The goal of this 
experiment is to determine whether 
the physics-based confinement method can correctly identify more native-like structures as having
low free energies.

@@

As is common in the CASP experiment, we assess our results in terms of Global Distance Test Total
Score (GDT-TS) \cite{Zemla2003}, which is a C$\alpha$ based measure of structural accuracy. It can
be understood roughly as the percentage of residues that are correctly positioned in the model
(range 0 to 100, higher is better). The initial models for the test was chosen from different server 
groups those have traditionally done well in past CASP events.

@@

Our next test was between models that are produced with different methodologies. 
Here, our model system was a fas apoptosis inhibitory protein molecule whose
pdb id and CASP target codes are 3MX7 and T0540 respectively. 
This protein contain 8 beta strands and 90 amino acid residues. 
The top models from groups "LTB" and "Mufold" was chosen for analysis.
We will call this as Model 1 and Model 2 for future references. As summerized in
Figure~\ref{fig:T0540}, the free energy based ranking once again correlates well with the 
GDT\_TS based score.
